---
title       : R programming
subtitle    : Lecture 6. Data manipulation
author      : Manuel A. Bolivar (Adapted from slides by Hadley Wickham)
job         : 
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : zenburn      # 
widgets     : []            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
---



```{r setup, include=FALSE}
# This is an R setup chunk, containing default options applied to all other chunks
library(knitr)
# This sets the chunk default options
opts_chunk$set(cache=TRUE, collapse=TRUE, error=FALSE, prompt=TRUE)
# This sets the chunk display theme
thm <- knit_theme$get("acid")
knit_theme$set(thm)
# This sets some display options
options(digits=3)
options(width=80)
```


<style>
em {
  font-style: italic
}
strong {
  font-weight: bold;
}
</style>

## Outline

1. `dplyr` package
2. One table verbs
3. Data pipelines


--- .class #id 

## `dplyr` package

+ Provides a flexible grammar of data manipulation.
+ Provide blazing fast performance for in-memory data by writing key pieces in C++.
+ Use the same interface to work with data no matter where it's stored, whether in a data frame, a data table or database.

```{r, results=F}
library(dplyr)
```

---

## `tbl_df()` function

+ Provides a compact view of a data frame

```{r}
data(mtcars)
cars <- tbl_df(mtcars)
cars
```

---

## One table verbs

+ **`filter`**: keep rows with matching condition.
+ **`select`**: select columns by name.
+ **`arrange`**: reorder rows.
+ **`mutate`**: add new variables.
+ **`summarise`**: reduce variables to values.

---

## Structure

+ First argument is a data frame.
+ Subsequent arguments say what to do with the data frame.
+ Always return a data frame.

---

## The `filter` verb

```{r}
df <- data.frame(
color = c("blue", "black", "blue", "blue", "black"),
value = 1:5)
df
```

```{r, eval = F}
filter(df,color == "blue")
```

---

## The `filter` verb

```{r}
df <- data.frame(
color = c("blue", "black", "blue", "blue", "black"),
value = 1:5)
df
```

```{r, eval = T}
filter(df,color == "blue")
```

---

## The `filter` verb

```{r}
df <- data.frame(
color = c("blue", "black", "blue", "blue", "black"),
value = 1:5)
df
```

```{r, eval = F}
filter(df,value %in% c(2,4))
```

--- .segue .dark .nobackground

## nycflights13 data

---


## nycflights13 data

```{r, warning=F}
library(nycflights13)
data(flights) #every flight departing NYC in 2013
dim(flights)
```
```{r}
data(weather)
dim(weather)
```

```{r}
data(planes)
dim(planes)
```


```{r}
data(airports)
dim(airports)
```

---

## flights data frame

```{r}
str(flights)
```

---

## Exercises

Find all flights:

+ To SFO or OAK
+ In January
+ Delayed more than an hour
+ That departed between midnight and five am.
+ That departed before 5am or after 10pm.

---

## The `select` verb

```{r}
select(df, color)
```

---

## The `select` verb

```{r}
select(df, color)
```

```{r}
select(df, -color)
```

---

## The `select` verb

Exist different ways to select columns using the `select` verb.

e.g. How to select the two delay variables in the flights data frame?

---

## The `select` verb

Exist different ways to select columns using the `select` verb.

e.g. How to select the two delay variables in the flights data frame?

```{r, eval = F}
select(flights, arr_delay, dep_delay)
select(flights, dep_delay, dep_delay + 2)
select(flights, ends_with("delay"))
select(flights, contains("delay"))
select(flights, one_of(c(arr_delay, dep_delay)))
select(flights, arr_delay:dep_delay) # Returns every thinkg between the vars
```

---

## The `arrange` verb

```{r}
arrange(df, color)
```

---

## The `arrange` verb

```{r}
arrange(df, color)
```

```{r}
arrange(df, desc(value))
```

---

## The `arrange` verb

```{r}
arrange(df, color, desc(value))
```

---

## Exercises

+ Order the flights by departure date and time.
+ Which flights were most delayed?
+ Which flights caught up the most time
+ during the flight?

---

## The `mutate` verb

```{r}
mutate(df, double = 2 * value)
```

---

## The `mutate` verb

```{r}
mutate(df, double = 2 * value,
       quadruple = 2*double)
```

---

## Exercises

1. Compute speed in mph from time (in minutes) and distance (in miles). Which flight flew the fastest?
2. Add a new variable that shows how much time was made up or lost in flight.
3. Create a new variable `date` that agregates year, month and day variables. Hint: try `ISOdate(year, month, day)`

---

## The `summarise` verb

```{r}
df
```


```{r}
summarise(df, total = sum(value))
```

---

## The `summarise` verb 

+ Most data operations are useful done on groups defined by variables in the the dataset.
+ The `group_by` function takes an existing tbl and converts it into a grouped tbl where operations are performed "by group".

---

## The `summarise` verb 

+ Most data operations are useful done on groups defined by variables in the the dataset.
+ The `group_by` function takes an existing tbl and converts it into a grouped tbl where operations are performed "by group".

```{r}
by_color <- group_by(df, color)
summarise(by_color, total = sum(value))
```

---

## The `summarise` verb 

```{r}
by_date <- group_by(flights, month, day)
by_hour <- group_by(flights, month, day, hour)
by_plane <- group_by(flights, tailnum)
by_dest <- group_by(flights, dest)
```

---

## Summary functions

+ `min(x)`, `max(x)`, `median(x)`, `mean(x)`
+ `n()`, `n_distinct()`, `sum(x)`
+ `sum(x>15)`, `mean(x>15)`

---

## Exercises

1. What is the plane with most flights along 2013?
2. Create a new data frame with the mean, median and standard deviation of departure delay by date.

--- .segue .dark .nobackground

## Data pipelines (%>%)

---

## Data pipelines (`%>%`)

The verbs and functions in R are really useful however it is hard t read multiple oprations.

```{r, results=F,echo=F}
flights <- mutate(flights,
                  date = ISOdate(year, month, day)
)
```

```{r}
hourly_delay <- filter(
  summarise(
    group_by(
      filter(
        flights,
        !is.na(dep_delay)
      ),
      date, hour
    ),
    delay = mean(dep_delay),
    n = n()
  ),
  n > 10
)
```

---

## Data pipelines (`%>%`)

**Solution**: The pipeline operator from `magrit` package

```{r}
hourly_delay <- flights %>%
  filter(!is.na(dep_delay)) %>%
  group_by(date, hour) %>%
  summarise(delay = mean(dep_delay), n = n())  %>%
  filter(n > 10)
```

---

## Exercises

Create data pipelines to answer the following
questions:

1. Which destinations have the highest average
delays?
2. Which flights operate every day of the year?

---

## References

+ RStudio. *Data wrangling with dplyr and tidyr*. https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf

