---
title       : R programming
subtitle    : Lecture 4. User-defined functions
author      : Manuel A. Bolivar
job         : 
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : zenburn      # 
widgets     : []            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
---


```{r setup, include=FALSE}
# This is an R setup chunk, containing default options applied to all other chunks
library(knitr)
# This sets the chunk default options
opts_chunk$set(cache=TRUE, collapse=TRUE, error=FALSE, prompt=TRUE)
# This sets the chunk display theme
thm <- knit_theme$get("acid")
knit_theme$set(thm)
# This sets some display options
options(digits=3)
options(width=80)
```


<style>
em {
  font-style: italic
}
strong {
  font-weight: bold;
}
</style>



## R functions

Functions are created using `fuction()` and are stored as R objects and are considered *first class objects*, i.e.,

  + Functions can be assigned to variables.
  + Functions can be stored in lists.
  + Functions can be nested (functions inside fuctions).
  + Functions can be passed as argumments of other functions.

--- .class #id 

## Syntaxis

The functions are R objects of class *function*.

```{r,eval = F, message=FALSE, warning=FALSE, include=TRUE}
f <- function(args){
  body #Do something interesting
}
```

where 
+ *args* is a set of symbol names.
+ *body* is an R expression.

---

## Arguments

Function arguments might include default values.
  + Arguments with default values are optional.
  + The *formal arguments* are the arguments included in the function definition.
  + The `args()` functions returns the function arguments.
  + A function call might not use all formal arguments.

---

## My first function

Desing a function that returns the mean of numeric vector.

---

## My first function

Desing a function that returns the mean of numeric vector.

```{r}
f <- function(x) {
  m <- sum(x)/length(x)
  return(m)
}
```

---

## My first function

Desing a function that returns the mean of numeric vector.

```{r}
f <- function(x) {
  m <- sum(x)/length(x)
  return(m)
}
```

```{r}
a <- c(2,4,6,8,10)
f(a)
```

---

## Argument matching

Arguments can be matched by position and name.

```{r}
args(plot) #argument names of plot function
```

All the following calls to the `plot()` function are equivalent.

```{r, eval= F}
plot(1:20,log(1:20))
plot(1:20,y = log(1:20))
plot(x = 1:20, y = log(1:20))
plot(y = log(1:20),x = 1:20)
```

---

## Return values

The `return()` function specifies the value (vector, list, data.frame, function) returned by the function. For example,

```{r}
f <- function(x){
  return(x^2+5)
}
```

However, if the `return()` function is not used, R returns the last expression in the *body* evaluated.

```{r}
f <- function(x){
  x^2+5
}
```

---

## Lazy evaluation

Arguments to functions are evaluated only as needed (*lazily*). For example,

```{r, error=T}
f <- function(a,b){
  a^3
}
```

```{r}
f(2) #Equivalent to f(a=2)
```

---

## Lazy evaluation

R executed the function's body statements sequentially.

```{r}
f <- function(a, b) {
print(a)
print(b)
}
```

```{r, error=T}
f(30)
```

---

## The `...` argument

The `...` arguments indicates a variable number of arguments. It is used when:
  + The number of arguments passed to the function is not known it advanced (e.g. `paste()`).
  ```{r}
  args(paste)
  ```
  + The functions is extending another function and uses the same arguments (e.g. `plot()`).
  ```{r}
  args(plot)
  ```
  + Take into acount that any argument after `...` must be named explicitly. 

---

## Exercises

1. Design a function that receive as argument a numeric vector and returns the descriptive statistics of the vector in a data frame. (`mean`, `sd`, `median`, `kurtosis`, `skewness`, `NAs`). 
Test your results using the `allmean_mv_EV` variable in the *COL PA MARA.csv* database.

2. Evenson`s cut points for PA are presented in the following table. The counts are presented using an epoch = 60 sec.

  ```{r,results="asis",echo=F}
  counts <- c("<=100","<=2292","<=4008",">4008")
  PA_Intensity <- c("sedentary","light","moderate","vigorous")
  evenson <- data.frame(counts,PA_Intensity)
  library(xtable)
  print(xtable(evenson),type="html")
  ```
  Create a function that receive as argument the count number in an epoch and returns the PA intensity level for that epoch.

---
  
## References

+ Adler, J. (2010). *R in a nutshell: A desktop quick reference*. O'Reilly Media, Inc.
+ Leek, J. (2014). *R programming (MOOC)*. Johns Hopkins University, Coursera.