---
title       : R programming
subtitle    : Lecture 5. R built-in functions
author      : Manuel A. Bolivar
job         : 
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : zenburn       # 
widgets     : []            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
---


```{r setup, include=FALSE}
# This is an R setup chunk, containing default options applied to all other chunks
library(knitr)
# This sets the chunk default options
opts_chunk$set(cache=TRUE, collapse=TRUE, error=FALSE, prompt=TRUE)
# This sets the chunk display theme
thm <- knit_theme$get("acid")
knit_theme$set(thm)
# This sets some display options
options(digits=3)
options(width=80)
```


<style>
em {
  font-style: italic
}
strong {
  font-weight: bold;
}
</style>

## Basic fuctions

+ `apply()`: Family of looping functions
+ `length()`: Return the number of elements in vector x.
+ `paste()` or `paste0()`: Concatenate vectors after converting to character.
+ `rep(1,5)`: Repeat the number 1 five times
+ `seq(1,10,0.4)`: Generate a sequence (1 -> 10, spaced by 0.4)
+ `sort()`: Sort the vector x
+ `order()`: List sorted element numbers of x
+ `summary()`: Returns a summary of a given object (e.g., data frame, lm, etc.).
+ `table()`: Returns a contigency table for each combination of factors.
+ `tolower()`, `toupper()`: Convert string to lower/upper case letters.
+ `unique()`: Remove duplicate entries from vector

--- .class #id 

## Basic fuctions

+ **`apply()`: Family of looping functions**
+ `length()`: Return the number of elements in vector x.
+ `paste()` or `paste0()`: Concatenate vectors after converting to character.
+ `rep(1,5)`: Repeat the number 1 five times
+ `seq(1,10,0.4)`: Generate a sequence (1 -> 10, spaced by 0.4)
+ `sort()`: Sort the vector x
+ `order()`: List sorted element numbers of x
+ `summary()`**: Returns a summary of a given object (e.g., data frame, lm, etc.).**
+ `table()`**: Returns a contigency table for each combination of factors.**
+ `tolower()`, `toupper()`: Convert string to lower/upper case letters.
+ `unique()`: Remove duplicate entries from vector.

---

## `table()`

The `table()` uses the cross-classifiying factors to build a contigency table of the count at each combination of factor levels.

---

## `table()`

The `table()` uses the cross-classifiying factors to build a contigency table of the count at each combination of factor levels.

```{r}
table(mtcars$cyl)
```

```{r}
table(mtcars$cyl,mtcars$am)
```

--- .segue .dark .nobackground

## apply family functions

---

## `apply` family functions

+ Using looping control structure as `for` and `while` are not that much efficient by themselfs. However, pre-allocating memory improves their performace.

+ `apply` functions use efficiently the resources by doing pre-allocation automatically.

+ Using `apply` functions make easy to parallelize your code.

---

## `apply` family functions

+ `lapply()`: To loop over a list and evaluate on each element.
+ `sapply()`: the same as `lapply`, but with the output in a more simpler form.
+ `apply`: To apply functions over array margins.
+ `tapply`: To apply a function to each cell of a ranged array. 


---

## `lapply()` function

```{r}
args(lapply)
```

`lapply` takes three arguments: 
  + `X` a list
  + `FUN` a function
  + `...` additional arguments of the function `FUN`

If `X` is not a list, it will be coerced to a list using `as.list()` function.

`lapply` returns a list.

---

## `lapply()` example

```{r}
lapply(mtcars,mean) #Returns the mean for each variable in a list
```

---

## `sapply()` function

```{r}
args(sapply)
```

`sapply` takes various arguments: 
  + `X` a list
  + `FUN` a function
  + `...` additional arguments of the function `FUN`
  + `simplify` a logical value. If `TRUE`, should return a simplified version of `lapply` function results. (e.g, vector, matrix or array) 

If `X` is not a list, it will be coerced to a list using `as.list()` function.

---

## `sapply()` example

```{r}
sapply(mtcars,mean) #Returns the mean for each variable in a vector
```

---

## Exercise 

1. What is the maximum value in each of the variables in `mtcars` data set?

---
## `apply()` function

```{r}
args(apply)
```

`apply` takes four arguments: 
  + `X` an array, including matrix
  + `MARGIN` a vector giving the subscripts which the function will be applied over. e.g., for a matrix 1 indicates rows, 2 indicates columns, c(1, 2) indicates rows and columns. 
  + `FUN` a function
  + `...` additional arguments of the function `FUN`

It is most often used to apply a function to the rows or columns of a matrix.

---

## `apply()` example

```{r}
x <- cbind(x1 = 3, x2 = c(4:1, 2:5))
x
```

---

## `apply()` example


```{r}
#col sums
apply(x, 2, sum) 
```

```{r}
#row sums
apply(x, 1, sum) 
```

---

## `tapply()` function

```{r}
args(tapply)
```

`apply` takes four arguments: 
  + `X` a vector (typically)
  + `INDEX` list of one or more factors, each of same length as `X`. The elements are coerced to factors by `as.factor()`
  + `FUN` a function
  + `...` additional arguments of the function `FUN`

Useful to summarise information across groups

---

## `tapply` example

```{r}
tapply(mtcars$wt, mtcars$cyl, mean)
```

```{r}
tapply(mtcars$wt, mtcars$cyl, sd)
```

---
## `tapply` example

```{r}
weightbycyl <- tapply(mtcars$wt, mtcars$cyl, mean)
barplot(weightbycyl)
```

---

## `tapply` example 2

```{r}
#tapply with multiple factors
tapply(mtcars$wt, list(mtcars$cyl,mtcars$carb), mean)
```

---

## Exercise 

1. Are, in average, cars with automatic transmission (`?mtcars`) heavier that those with manual transmission?
2. Compare `tapply()` and `table()` functions. How are they similar? How are they different?


--- .segue .dark .nobackground

## summary function

---

## `summary()`

`summary` is a generic function used to produce result summaries of the results of various model fitting functions.

```{r}
data(mtcars) #Load mtcars data to environment
summary(mtcars)
```

---

## Exercises

1. Design a function that receive as argument a numeric vector and returns the descriptive statistics of the vector in a data frame. (`mean`, `sd`, `median`, `minimum`, and `maximum`)
2. Report the descriptive statistics for all the variables in `mtcars` using only one code line.
3. Report the descriptive statistics for the weigth variable (`wt`) group by the transmission type.(`?mtcars`). Use only one code line.

---

## Homework

**PENDING**

---
## References

+ Koduvely, D. H. M. (2015). *Learning Bayesian Models with R*. Packt Publishing Limited.
+ Ross, N. (2013). *FasteR! HigheR! StrongeR! - A Guide to Speeding Up R Code for Busy People*. http://www.noamross.net/blog/2013/4/25/faster-talk.html
