---
title       : R programming
subtitle    : Lecture 2. Subsetting
author      : Manuel A. Bolivar
job         : 
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : zenburn       # tomorrow
widgets     : []            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
---

```{r setup, include=FALSE}
# This is an R setup chunk, containing default options applied to all other chunks
library(knitr)
# This sets the chunk default options
opts_chunk$set(cache=TRUE, collapse=TRUE, error=FALSE, prompt=TRUE)
# This sets the chunk display theme
thm <- knit_theme$get("acid")
knit_theme$set(thm)
# This sets some display options
options(digits=3)
options(width=80)
```

<style>
em {
  font-style: italic
}
strong {
  font-weight: bold;
}
</style>

## Subsetting

+ Subsetting allows you to pull out the pieces that you’re interested in. 
+ R’s subsetting operators are powerful and fast.
+ Subsetting can be applied to vectors, lists, factors, arrays, and data frames. 


--- .class #id 

## Vectors

```{r}
x <- c(2.1, 4.2, 3.3, 5.4)
```

`Positive integers:` Return elements at the specified positions.

---

## Vectors

```{r}
x <- c(2.1, 4.2, 3.3, 5.4)
```

`Positive integers:` Return elements at the specified positions.

```{r}
x[c(2,4)]
```

```{r}
x[order(x)]
```

```{r}
x[c(1,1)]
```

---

## Vectors

```{r}
x <- c(2.1, 4.2, 3.3, 5.4)
```

`Negative integers:` omit elements at the specified positions.

---

## Vectors

```{r}
x <- c(2.1, 4.2, 3.3, 5.4)
```

`Negative integers:` omit elements at the specified positions.

```{r}
x[-c(2,4)]
```

but you can´t mix positive and neative integers.

```{r,error=TRUE}
x[c(-2,4)]
```


---

## Vectors

```{r}
x <- c(2.1, 4.2, 3.3, 5.4)
```

`Logical vectors`: select elements where the corresponding logical value is `TRUE`.

```{r}
x[c(TRUE,FALSE,TRUE,TRUE)]
```

```{r}
x[x>3]
```

---

## Vectors

```{r}
x <- c(2.1, 4.2, 3.3, 5.4)
```

`Logical vectors`: select elements where the corresponding logical value is `TRUE`.


---

## Vectors

```{r}
x <- c(2.1, 4.2, 3.3, 5.4)
```

`Logical vectors`: select elements where the corresponding logical value is `TRUE`.

```{r}
x[c(TRUE,FALSE,TRUE,TRUE)]
```

```{r}
x[x>3]
```

If the logical vector is shorter than the vector being subsetted, it will be recycled to be the same length.

```{r}
x[c(TRUE, FALSE)]
x[c(TRUE, FALSE,TRUE, FALSE)]
```

---

## Matrices and arrays

The most common way of subsetting matrices (2d) and arrays (>2d) is a simple generalisation of 1d subsetting: 

+ Supply a 1d index for each dimension, separated by a comma. 
+ Blank subsetting is now useful because it lets you keep all rows or all columns.

```{r}
a <- matrix(1:9, nrow = 3)
a
```

---

## Matrices and arrays - examples

```{r}
a
```

Write the statements that return each of the following statements for matrix `a`:

 + 3th row
 + 2nd column
 + 1st and 2nd column
 + Element located at the 2nd row and second column
 + The intersection between the 2nd and 3rd rows and the 1st and 3rd columns

---

## Matrices and arrays - examples

```{r}
a
```

```{r}
a[,3] # 3th  column of the matrix
```

```{r}
a[2,] # 2th row of the matrix
```

```{r}
a[c(2,3),c(1,3)] # 2nd and 3rd rows and 1st and 3rd columns of the matrix
```

---

## Data frames

+ Data frame subsetting works like matrices 
+ Now we have an additional operator `$` besides `[`.

```{r}
df <- data.frame(ID = 1:4, 
                 Color = c("red","white","red",NA),
                 Passed = c(T,T,T,F),
                 stringsAsFactors = F)
df
```

--- &twocol

## Data frames - examples

Using the `[`, we will obtain a data frame.

```{r}
str(df)
```

*** =left

```{r}
df[1]
class(df[1])
```

*** =right

```{r}
df["ID"]
class(df["ID"])
```

--- &twocol

## Data frames - examples

Using the `[[` or `$`, we obtain a vector.

```{r}
str(df)
```

*** =left

```{r}
df$ID
class(df$ID)
```

*** =right

```{r}
df[["ID"]]
class(df[["ID"]])
```


---

## Lists

Subseting lists works the same than subsetting a vector.

+ `[` returns a list.
+ `[[` and `$` return the components of the list.

```{r}
x <- list(a=c(5:9),b=c(TRUE,FALSE,FALSE,TRUE), c=list("four","five"))
str(x)
```

---

## Lists

```{r}
x <- list(a=c(5:9),b=c(TRUE,FALSE,FALSE,TRUE), c=list("four","five"))
```

For example, `x[1]` and `x["a"]` return a list.

```{r}
x[1]
class(x[1])
```

```{r}
x["a"]
class(x["a"])
```

---

## Lists

```{r}
x <- list(a=c(5:9),b=c(TRUE,FALSE,FALSE,TRUE), c=list("four","five"))
```

But `x$a`, `x[[1]]`, and `x[["a"]]` return a vector.

```{r}
x[[1]]
class(x[[1]])
```

```{r}
x$a
class(x$a)
```

---

## Exercises

1. If a matrix `a` is defined as `a <- matrix(1:9, nrow = 3)`, what would return `a[c(TRUE,TRUE,FALSE),c(-3)]`?
2. Fix each of the following common data frame subsetting errors:
  + `mtcars[mtcars$cyl = 4, ]`
  + `mtcars[-1:4, ]`
  + `mtcars[mtcars$cyl <= 5]`
  + `mtcars[mtcars$cyl == 4 | 6, ]`
  
3. Given a linear model, e.g., `mod <- lm(mpg ~ wt, data = mtcars)`, extract the residual degrees of freedom. Extract the R squared from the model summary (`summary(mod)`)

---

## References

+ Wickham, H. (2014). *Advanced R*. CRC Press.