---
title       : R programming
subtitle    : Lecture 3. Control structures
author      : Manuel A. Bolivar
job         : 
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : zenburn      # 
widgets     : []            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
---

```{r setup, include=FALSE}
# This is an R setup chunk, containing default options applied to all other chunks
library(knitr)
# This sets the chunk default options
opts_chunk$set(cache=TRUE, collapse=TRUE, error=FALSE, prompt=TRUE)
# This sets the chunk display theme
thm <- knit_theme$get("acid")
knit_theme$set(thm)
# This sets some display options
options(digits=3)
options(width=80)
```


<style>
em {
  font-style: italic
}
strong {
  font-weight: bold;
}
</style>


## Motivation

+ The statements in R are executed sequentially (normally).
+ But sometimes you might:
  - Execute some statemets repetitively (`for`, `while`)
  - Execute statements only if certain conditions are met (`if-else`)
  
**e.g.**
  + Read multiple files at once (.agd).
  + Identify valid participants given their MVPA minutes.

--- .segue .dark .nobackground

## Repetition and looping

--- .class #id 



## Repetition and looping

Looping constructs repetitively execute a statement or series of statements until a condition is not true.

 + `for`
 + `while`
 + `repeat`*

---

## For

The `for` loop executes a statement repetitively until a variable’s value is no longer contained in the sequence `seq`. The syntaxis is:

```{r, eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
for(var in seq){
  statement
}
```

where,
+ *var* is a variable name,
+ *seq* is a vector of any type, and
+ *statement* is the statement you want to execute repetitively.

---

## For - Example 1

Print the numbers from 1 to 10
```{r}
nums <- 1:10
```

```{r}
  for(i in nums){
    print(i)
  }
```


---

## For - Example 2

The for loop can also iterate other data type vectors.

```{r}
  weekdays <- c("Monday","Tuesday","Wednesday","Thursday","Friday")
```

```{r}
  for(day in weekdays){
    print(day)
  }
```

---

## For - Example 3

`seq_along(x)` function returns a numeric vector of the form `1:length(x)`. It can be used as argument for the for loop.

```{r}
  weekdays <- c("Monday","Tuesday","Wednesday","Thursday","Friday")
```

```{r}
  for(i in seq_along(weekdays)){
    day <- weekdays[i]
    print(day)
  }
```

---

## For - Exercises

1. Write a program which sums the integers from 1 to 10 using a `for` loop (and prints the total at the end).
2. Write a program which finds the factorial of a given number. E.g. 3 factorial, or 3! is equal to 3 x 2 x 1; 5! is equal to 5 x 4 x 3 x 2 x 1, etc.. Your program should only contain a single loop.

---

## While

A `while` loop executes a statement repetitively until the condition is no longer true. The sintaxys is: 

```{r,eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
while(cond){
  statement
}
```

where,
  + *cond* is the condition that should be TRUE for keep iterating in the loop.

---

## While - Example 1

Add the numbers from 1 to 10 together.

```{r}
sum <- 0
i <- 1
while(i<=10){
  sum <- i + sum
  i <- i + 1
}
```


```{r}
sum
```

---

## While - Exercises

1. Write a program which uses a while loop to sum the squares of integers (starting from 1) until the total exceeds 200. Print the final total and the last number to be squared and added.

--- .segue .dark .nobackground

## Conditional execution

---

## Conditional execution

In conditional execution, a statement or statements are executed only if a specified condition is met. 

These constructs include:
  + `if-else`
  + `ifelse`

---

## if-else

The `if-else` control structure executes a statement if a given condition is true.

```{r,eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
# Option 1
if(cond){
  statement #do something
}
```

```{r,eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
# Option 2
if(cond){
  statement1 # do something
} else{
  statement2 # do somenthing else
}
```

---

## if-else

```{r,eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
# Option 3
if(cond1){
  statement1 # do something
} else if(cond2){
  statement2 # do somenthing different
} else{
  statement3 # do something else
}
```

---

## if-else

The conditions can be a compound conditions, e.g.
  + `if(cond1 & cond2) {statement}`: `&` is an *and* operator. 
  + `if(cond1 | cond2) {statement}`: `|` is an *or* operator. 

---

## if-else - Example 1

A participant with more than 600 minutes of moderate-to-vigorous physical activity (MVPA) in average is considered valid.

```{r}
minutes <- 750
if(minutes > 600){
  print("Valid participant")
}
```

---

## if-else - Example 2

A participant with more than 600 minutes of moderate-to-vigorous physical activity (MVPA) in average is considered valid. **Otherwise, they are invalid participants.**

```{r}
minutes <- 450
if(minutes > 600){
  print("Valid participant")
} else{
  print("Invalid participant")
}
```

---

## if-else - Example 3

A participant with more than 600 minutes of moderate-to-vigorous physical activity (MVPA) in average **and more than 4 valid days of wear time** is considered valid. Otherwise, they are invalid participants.

```{r}
minutes <- 750
valdays <- 3
if(minutes > 600 & valdays > 4){
  print("Valid participant")
} else{
  print("Invalid participant")
}
```


---

## Logical operators

```{r, results='asis',echo=F}
library(xtable)
tab <- data.frame(Operator=c('<','<=','>','>=','==','!=','!x','x | y','x & y'),Description=c('less than
','less than or equal to','greater than','greater than or equal to','exactly equal to','not equal to','Not x','x OR y','x AND y'))
print(xtable(tab), type = "html",include.rownames = F,size=2)
```

---

## if-else - Exercises

1. Run the following code chunk.
  ```{r,eval=FALSE}
  set.seed(123) #Set the seed for the random number generation
  x <- rnorm(n = 1000, mean = 1, sd = 3)
  ```
  + How many numbers in `x` are less or equal than 0. Use a `for` or `while` loop.
  + How many numbers in `x` are more than 0 and less or equal than 1. Use a `for` or `while` loop.



---

## ifelse

The `ifelse` construct is a compact and vectorized version of the `if-else` construct.
The syntax is:

```{r,eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
ifelse(cond,statement1,statement2)
```

+ The first statement is executed if *cond* is `TRUE`. If *cond* is `FALSE`, the second statement is executed.

+ Use `ifelse` when you want to take a binary action or when you want to input and output vectors from the construct.

---

## Exercises

1. Read the database *COL PA MARA.csv*. ([data](https://goo.gl/77zASV)) ([codebook](https://goo.gl/28MBXi))
  ```{r, eval= FALSE}
  acdata <- read.csv("COL PA MARA.csv")
  ```
  A valid participant is the one who wear the accelerometer for more than 4 valid days and at least one valid weekend day. However, if a person wear the accelerometry for more than 2000 minutes in midweek, he is also considered a valid participant.
  + Create a new logical variable named `valid` which indicates if a particiant is valid or not. Bind your new variable to `acdata` data frame. (Use loops and conditional statements)
  + How many participants are valid?
  
2. Using the `ifelse()` function, create a new logical variable named `valid` which indicates if a particiant is valid or not.
  + Compare the code used in numerals 1 and 2, which one of them is more efficient?







---

## Remarks

+ Looping in R can be inefficient and time consuming when you’re processing the rows or columns of large datasets.
+ Whenever possible, it’s better to use R’s built-in numerical and character functions in conjunction with the apply family of functions.

---

## References

+ Kabacoff, R. (2015). *R in action: data analysis and graphics with R*. Manning Publications Co.

