---
title       : R programming
subtitle    : Lecture 3. Control structures
author      : Manuel A. Bolívar
job         : 
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : zenburn      # 
widgets     : []            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
---

```{r setup, include=FALSE}
# This is an R setup chunk, containing default options applied to all other chunks
library(knitr)
# This sets the chunk default options
opts_chunk$set(cache=TRUE, collapse=TRUE, error=FALSE, prompt=TRUE)
# This sets the chunk display theme
thm <- knit_theme$get("acid")
knit_theme$set(thm)
# This sets some display options
options(digits=3)
options(width=80)
```


<style>
em {
  font-style: italic
}
strong {
  font-weight: bold;
}
</style>


## Motivation

+ The statements in R are executed sequentially (normally).
+ But sometimes you might:
  - Execute some statemets repetitively (`for`, `while`)
  - Execute statements only if certain conditions are met (`if-else`)
  
**e.g.**
  + Read multiple files at once (.agd).
  + Identify valid participants given their MVPA minutes.

--- .segue .dark .nobackground

## Repetition and looping

--- .class #id 



## Repetition and looping

Looping constructs repetitively execute a statement or series of statements until a condition is not true.

 + `for`
 + `while`
 + `repeat`*

---

## For

The `for` loop executes a statement repetitively until a variable’s value is no longer contained in the sequence `seq`. The syntaxis is:

```{r, eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
for(var in seq){
  statement
}
```

---
## While

A `while` loop executes a statement repetitively until the condition is no longer true. The sintaxys is: 

```{r,eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
while(cond){
  statement
}
```

--- .segue .dark .nobackground

## Conditional execution

---

## Conditional execution

In conditional execution, a statement or statements are executed only if a specified condition is met. 

These constructs include:
  + `if-else`
  + `ifelse`

---

## if-else

The `if-else` control structure executes a statement if a given condition is true.

```{r,eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
# Option 1
if(cond){
  statement #do something
}
```

```{r,eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
# Option 2
if(cond){
  statement1 # do something
} else{
  statement2 # do somenthing else
}
```

---

## if-else

```{r,eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
# Option 3
if(cond1){
  statement1 # do something
} else if(cond2){
  statement2 # do somenthing different
} else{
  statement3 # do something else
}
```


---

## ifelse

The `ifelse` construct is a compact and vectorized version of the `if-else` construct.
The syntax is:

```{r,eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
ifelse(cond,statement1,statement2)
```

+ The first statement is executed if *cond* is `TRUE`. If *cond* is `FALSE`, the second statement is executed.

+ Use `ifelse` when you want to take a binary action or when you want to input and output
vectors from the construct.

---

## Exercises

1. 
2. 
3.






---

## Remarks

+ Looping in R can be inefficient and time consuming when you’re processing the rows or columns of large datasets.
+ Whenever possible, it’s better to use R’s built-in numerical and character functions in conjunction with the apply family of functions.

---

## References

+ Kabacoff, R. (2015). *R in action: data analysis and graphics with R*. Manning Publications Co.

--- .segue .dark .nobackground

## Birthday Problem